When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-02-13 doc/paper/vpd_et_paper.tex
some progress on pierre's comments, still thinking of delta role though

Differences in $VPD_{crit}$ are exclusively determined by $uWUE$ and the slope parameter ($g_1$) related to the plant functional type. A larger underlying water use efficiency, $uWUE$, causes means that at the ecosystem scale $WUE$ decreases more with increasing VPD, so plants lose more water per unit carbon production and are more likely to increase ET with increasing VPD (REFLRMUALTE SENTENCE, THE RESULTS IS CONTER INUITTICE WE EDPECT THE OPPOSITE). A tendency towards increasing ET response with increasing VPD can also be caused by a high slope parameter ($g_1$), characteristic of plants that at the leaf scale are more willing to trade water for stomata opening. Plants that are less conservative will be thus be more likely to increase ET with increasing VPD. Both the aforementioned effects (large $uWUE$, $g_1$) can amplify each other, and generally conspire to shift the sign term towards a positive value for a given PFT.

This effect of $uWUE$ and $g1$ on the sign term is most apparent by comparing two extreme PFTs: water intensive CRO and water conservative CSH. CRO has a high slope parameter ($g_1 = 183.1 \, Pa^{1/2}$) compared to CSH ($g1 = 148.6 \, Pa^{1/2}$. It also has a slightly higher uWUE ($2.32$ -units-) compared to CSH ($uWUE=2.11$ -units-). MAYBE A DUMB QUESTION BUT CAN'T uWUE AND G1 BE ANALYTICALLY RELTED> I THINK AT LESAT YOU SHOULD MAKE CLEAR THAT UWUE IS AT THE ECOYSTEM SCALE (DID YOU FIT IT OR USED PUBLISHED DATA-I AM NOW CONFUSED). These differences in slope parameter cause opposite ET responses to changes in VPD between CRO and CSH. ET theoretically always increases with increasing VPD for the more water intensive CRO, while ET almost always decreases with increasing VPD for the more water conservative CSH (Figure \ref{term3}). CRO, evolved or were bred to prioritize GPP and yield and thus not water conservative so that they are very willing to trade water for photosynthesis and productivity, despite changes in VPD, while CSH are very unwilling to trade water for more photosynthesis.

AS expected, the slope parameter ($g_1$) is a primary determinant of the VPD dependence for the sign term shown in Figure \ref{term3}. Plants that are more conservative (small $g_1$) will tend to reduce ET with increasing VPD, and will be very effective at reducing ET, especially at low VPD. However, at very high VPD, gradients in vapor pressure at the leaf scale will be so strong that they will begin to dominate leaf-scale plant strategies of stomatal (partial) closure (parameterized with $g_1$). As a result, ET response will begin to asymptote towards its ecosystem-scale values irregardless of leaf-scale strategy and stomatal regulation (parameterized with $g_1$). Therefore, plants with a low $g_1$ will have the largest VPD dependence of ET response because the difference in ET response at low VPD (leaf stomatal response dominates) and high VPD (VPD gradient dominates) is largest. This is apparent in the strong VPD dependence of ENF, which has the lowest slope parameter ($g_1=74.31$), in Figure \ref{term3}.

@@

$\frac{P}{T}$ is an air-density term, which we varies little compared to aerodynamic conductance and Clausius-Clapeyron ($\Delta$). The psychometric constant ($\gamma$) is also relatively constant, so the scaling term should be primarily a function of aerodynamic conductance and temperature, through the Clausius-Clapeyron relationship $\Delta$. This is as expected, given that the aerodynamic conductance represents the efficiency of exchange between the surface and the atmosphere. As aerodynamic conductance  increases, any plant response will be communicated more strongly to the atmosphere (and vice-versa).

>>>>

Differences in $VPD_{crit}$ are exclusively determined by $uWUE$ and the slope parameter ($g_1$) related to the plant functional type. A larger $uWUE$ means a smaller $VPD_{crit}$, and an ET response to increases VPD that is more likely to be positive. At first glance this result is somewhat counterintuitive; we expect that plants with a higher water use effiency would be more water conservative. However, in reality $uWUE$ determines how $WUE$ changes with VPD:

\[WUE = \frac{GPP}{ET} = \frac{uWUE}{\sqrt{VPD}}.\]
\[\frac{\partial \; WUE}{\partial \; VPD} = -\frac{uWUE}{2 \; VPD^{3/2}}\]

So, plants with a higher $uWUE$ will have a greater \textit{decrease} in ecosystem-scale $WUE$ in response to increases in VPD. This decrease in $WUE$ causes more water loss per unit carbon production, and explains the relationship between high $uWUE$ and high liklihood of increases of ET in response to increasing atmospheric drying (increases in VPD).

A tendency towards increasing ET response with increasing VPD can also be caused by a high slope parameter ($g_1$), characteristic of plants that at the leaf scale are more willing to trade water for stomata opening. Plants that are less conservative will be thus be more likely to increase ET with increasing VPD. Both the aforementioned effects (large $uWUE$, $g_1$) can amplify each other, and generally conspire to shift the sign term towards a positive value for a given PFT.

This effect of $uWUE$ and $g1$ on the sign term is most apparent by comparing two extreme PFTs: water intensive CRO and water conservative CSH. CRO has a high slope parameter ($g_1 = 183.1 \, Pa^{1/2}$) compared to CSH ($g1 = 148.6 \, Pa^{1/2}$. It also has a slightly higher uWUE ($2.32$ -units-) compared to CSH ($uWUE=2.11$ -units-). \explain{MAYBE A DUMB QUESTION BUT CAN'T uWUE AND G1 BE ANALYTICALLY RELTED> I THINK AT LESAT YOU SHOULD MAKE CLEAR THAT UWUE IS AT THE ECOYSTEM SCALE (DID YOU FIT IT OR USED PUBLISHED DATA-I AM NOW CONFUSED)}. These differences in slope parameter cause opposite ET responses to changes in VPD between CRO and CSH. ET theoretically always increases with increasing VPD for the more water intensive CRO, while ET almost always decreases with increasing VPD for the more water conservative CSH (Figure \ref{term3}). CRO evolved or were bred to prioritize GPP and yield and are thus not water conservative. They are very willing to trade water for photosynthesis and productivity, despite changes in VPD, while CSH are very unwilling to trade water for more photosynthesis.

As expected, the slope parameter ($g_1$) is a primary determinant of the VPD dependence for the sign term shown in Figure \ref{term3}. Plants that are more conservative (small $g_1$) will tend to reduce ET with increasing VPD, and will be very effective at reducing ET, especially at low VPD. However, at very high VPD, gradients in vapor pressure at the leaf scale will be so strong that they will begin to dominate leaf-scale plant strategies of stomatal (partial) closure (parameterized with $g_1$). As a result, ET response will begin to asymptote towards its ecosystem-scale values irregardless of leaf-scale strategy and stomatal regulation (parameterized with $g_1$). Therefore, plants with a low $g_1$ will have the largest VPD dependence of ET response because the difference in ET response at low VPD (leaf stomatal response dominates) and high VPD (VPD gradient dominates) is largest. This is apparent in the strong VPD dependence of ENF, which has the lowest slope parameter ($g_1=74.31$), in Figure \ref{term3}.

@@

$\frac{P}{T}$ is an air-density term, which varies little compared to aerodynamic conductance and Clausius-Clapeyron ($\Delta$). The psychometric constant ($\gamma$) is also relatively constant, so the scaling term should be primarily a function of aerodynamic conductance and temperature, through the Clausius-Clapeyron relationship $\Delta$. This is as expected, given that the aerodynamic conductance represents the efficiency of exchange between the surface and the atmosphere. As aerodynamic conductance  increases, any plant response will be communicated more strongly to the atmosphere (and vice-versa).
